---
title: "Homework 3 - Task 3"
author: "Carmen Galaz-García"
date: "11/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)  # all code chunks message=FALSE

library(tidyverse)   # data wrangling
library(here)        # reference folders withing project
library(janitor)     # clean data
library(kableExtra)  # create nice tables
library(lubridate)   # wrangle dates

urchins <- read.csv(here("data","urchins.csv"))  # read data
urchins <- clean_names(urchins)                  # simplify data names
```


# 1. Size comparisons between purple and red urchins


## Exploratory data visualization 
The following graphs compare the distribution of urchin diameters (in cm) for purple urchin (*Strongylocentrotus purpuratus*) and red urchin (*Mesocentrotus franciscanus*) observed at 5 sites along the Santa Barbara coast.

### Histograms 


```{r}
ggplot(data = urchins, aes(x=size)) +   # select data and variable
  geom_histogram()+                     # do histogram
  facet_wrap(~common_name, scales = "free_x")+ # facet according to common_name and choose each x-axis length individually
  labs(x="Urchin diameter (cm)", y ="Count")   # update labels
```

### Boxplots
```{r}
ggplot(data=urchins, aes(x=common_name, y=size))+ # select data and variables
  geom_boxplot()+        # do boxplot
  labs( x="Common name", # update labels
        y="Urchin diameter (cm)")
```

## Descriptive statistics table
The table below contains the following statistics for purple and red urchins (ignoring site as a variable): mean, median, standard deviation, and sample size.

```{r}
urchins_stats <- urchins %>%           # copy all urchins data
  group_by(common_name) %>%            # analyze each group separately
  summarize(mean_size = mean(size),    # for each group calculate: size mean
            meadian_size=median(size),      # size median
            sd_size = sd(size),             # size standard deviation
            n =n())                         # number of individuals

# create table using kable
urchins_stats %>% kable( col.names = c("Common name", 
                                       "Mean diameter (cm)", 
                                       "Median size (cm)",
                                       "Standard deviation (cm)",
                                       "Sample size (n)" ) 
                         ) %>% 
   kable_styling(full_width = FALSE)
```

### Analysis

We can see in the histogram for purple urchins that their size is slightly positively skewed. In their boxplot we can further see that there is a long tail with possible outliers which have big diameter. 

For red urchins we can see  there is a bit of a tail to the right but otherwise the distribution looks quite normal. This is also apparent in the boxplot showing more outliers with big diameter but otherwise a fairly symmetric box and whiskers. 
 
We can see too that on average red urchins have a diameter (6.776489-4.469797)/4.469797* 100=51.6% bigger than that of a purple urchin. The absolute differences between the standard deviations of both populations is 0.7 cm, with red urchins presenting a greater variability, which is (1.891343-1.175264)/1.175264* 100 = 60.9% of that of purple urchins. 

# 2. Exploring purple urchin counts

### Wrangling data
Using a piped sequence, we create a subset of the urchins data which only includes observations for purple urchins, has dates in class "Date", has a column wich only contains the year of the observation and adds a column with the full name of the observation site. 

```{r}
purple_urchins <- urchins %>%                   # assign dataset
  filter(common_name == "Purple Urchin") %>%    # select rows corresponding to purple urchins
  mutate(Year = year(mdy(date))) %>%            # create Year column 
  mutate(site_full = case_when( site == "NAPL" ~ "Naples Reef",          # create column with complete names
                                site == "AQUE" ~ "Arroyo Quemado Reef",
                                site == "CARP" ~ "Carpinteria Reef",
                                site == "MOHK" ~ "Mohawk Reed",
                                site == "IVEE" ~ "Isla Vista Reef"))
```
 


### Summary table
The table below shows the total number of purple urchins observed by site and year.

```{r}
purples_summary <- purple_urchins %>%  # starting with purple_urchins data 
  count(site_full, Year)               # count # individuals in each site in a given year

# create table
purples_summary %>% kable(col.names = c("Site", "Year", "Sample size (n)")) %>% 
  kable_styling(full_width = FALSE)

```

### Exploratory graph
In the following graphs we see the change in the counts of purple urchins recorded at each site over time for the years included in the study.

```{r}
ggplot( data = purples_summary, aes(x=Year, y=n))+        # select data and variables
  geom_line() +                                           # line+point to join the dots
  geom_point() +
  facet_wrap(~site_full, scales = "free_x") +             # facet by site
  scale_x_continuous(breaks=2*(1004:1010))  +             # update x-axis ticks
  labs(y="Number of urchins at site")

```











