---
title: "Homework 3 - Task 4"
author: "Carmen Galaz-García"
date: "11/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)  # data wrangling
library(janitor)    # clean names
library(lubridate)  # date wrangling
library(here)       # folder location 
library(kableExtra) # nice tables

urchins <- read_csv(here("data","urchins.csv"))
urchins <- clean_names(urchins)    # change all-caps column names
```

## Data wrangling

Selection of data corresponding to Naples Reef observations for purple urchins observed in January 2010 


```{r}
NPL_purple_Jan2010 <- urchins %>%    # copy data frame
  mutate(new_date = mdy(date)) %>%   # create column with dates in Date class
  select(!date) %>%                  # remove column with old dates
  filter(site == "NAPL" &            # select Naples Reef site 
           common_name == "Purple Urchin" &    # select Purple Urchins
           month(new_date)==01 &     # select month = January 
           year(new_date)==2010)     # select year = 2010
```


## Exploratory Data Visualization

Histograms and QQ-plots for the size distribution for each of the treatments (Control and Annual) to assess normality. 

```{r}
ggplot(data= NPL_purple_Jan2010, aes(x=size))+
  geom_histogram(bins = 10) +     # make histogram, update number of bins
  facet_wrap(~treatment) +        # create one plot per treatment
  labs(x="urchin diameter (cm)")  # updating x-axis label
```

```{r}
ggplot(data = NPL_purple_Jan2010, aes(sample=size))+
  geom_qq()+
  facet_wrap(~treatment)
```

Looking at the QQ-plots we can say it is probable that the size distribution for both the anual and control groups are normally distributed, since the plot roughly follows the path of a straight line. However looking at the histograms we can see there is a slight bimodal behaviour in both groups, also present in the QQ-plots as there are are two fairly long horizontal accumulations of points in each group.


## Descriptive statistics

Table of descriptive statistics for both treatment groups
including the mean, standard deviation, and sample size.    
```{r}
urchin_stats <- group_by(NPL_purple_Jan2010, treatment) %>%  # group previously selected data by treatment
  summarize(mean_size = mean(size),    # get mean size
            meadian_size=median(size), # get median size
            sd_size = sd(size),        # get standard deviation
            n =n())                    # get sample size

urchin_stats %>% kable( col.names = c( "Treatment",            # column headers 
                                       "Mean diameter (cm)", 
                                       "Median size (cm)",
                                       "Standard deviation (cm)",
                                       "Sample size (n)" ) 
                         ) %>% 
   kable_styling(full_width = FALSE)
```

## Two-sample t-test

In this section we perform a two-sample t-test to compare means between the two groups.

```{r}
# select size vector of annual treatment urchins
annual_sample <- filter(NPL_purple_Jan2010, treatment == "ANNUAL") %>% 
  pull(size) #extracts size columns as a vector

# select size vector of no treatment (control) urchins
control_sample <-filter(NPL_purple_Jan2010, treatment == "CONTROL") %>% 
  pull(size)

# two-sample t-test
t.test(x=annual_sample, y=control_sample) 
```
Urchins coming from sites with annual treatment have a mean diameter length of 3.92 cm (rounded to two decimals) while urchins from sites with no treatment have a mean diameter length of 3.73 cm, the absolute difference between these two is 0.19cm. We can see too that annual treatment urchins are on average (3.916667-3.725806)/3.725806*100 = 5.122% bigger than no treatment urchins. On the other hand a two-sample t-test for these population gives a p-value of 0.2536, which indicates there is at least a 1/4 probability of drawing two samples with means having a difference of at least 0.19cm if they come from populations with the same mean. So we could say there is quite a big chance that these two samples come from populations that actually have the same mean, from which we could conclude that annual treatment probably does not affect the urchin’s size.


